### 2023-08-15 Manuscript/Draft2.Rmd
manuscript updates

# Estimating Population Trends with Stratified Random Sampling Under the Pressures of Climate Change

@@

Corresponding author: Ben Levy (benjamin.levy@noaa.gov)

@@

Many stock assessments in the United States use abundance estimates derived from stratified random bottom trawl data. To accurately track changes in true abundance of a species, catch observations must contain low noise and the proportion of the population that is sampled should be consistent over time. One increasingly documented phenomenon that violates the latter assumption is climate-induces changes in relative abundance between sampled and unsampled portions of the stock area. Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to evaluate two methods of estimating population trend given temperature driven movement. Movement rates combine species-specific static habitat preferences with temperature tolerances. Habitat preferences were derived from niche models relating bottom trawl catches to environmental covariates. A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation. We simulated spatial time series datasets for each species for several temperature scenarios and population trends. Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST). Our focus was on the ability of contemporary index standardization methods to track population trends under shifting spatial preferences.

<!-- We found that in general, both index calculation methods are able to produce similar abundance trends to the model output, with model-based estimates providing more consistent year-to-year estimates. The design-based approach tended to provide abundance estimates that were below the true values in our models, while model-based estimates tended to overestimate true abundance. We highlight scenarios where index calculation methods produced misleading estimates as species shifted their spatial densities over time within their assumed stock boundaries. While including environmental covariates can improve model-based estimates, the outcome is sensitive to the multitude of inputs and settings. The impact of shifting spatial densities became magnified when only a portion of a species’ spatial distribution was sampled. Counter-intuitively, including covariates can sometimes degrade the estimates in this situation. -->

@@

The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf, where depth and latitude define the strata. The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each stratum in approximately the same 6 week time period in each season.

One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process. Common model-based approaches include Generalized Linear Models (GLM), Generalized Additive Model (GAM), and General Linear Mixed Models (GLMM).

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. At the same time a number of recent studies have found a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects]. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate]. A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence]. In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences]. We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.

@@

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depends on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends. We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance]. VAST also allows users to include covariate data to better inform the model. Covariates can be static (e.g. habitat preferences) or dynamics (e.g. temperature). We explore whether including environmental predictors leads to improved abundance estimates, which is particularly relevant as climate change progresses.

@@

We construct spatial population models for Yellowtail Flounder (*Limanda ferruginea*), Atlantic Cod (*Gadus morhua*), and Haddock (*Melanogrammus aeglefinus*) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock. Movement of each species combines static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50**??** years while Haddock has increased in abundance in recent years [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **just cite 2021 assessments?**. For this reason we compare survey index estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. We consider increasing, decreasing, and constant population trends for Yellowtail for completeness. The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios). Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited stratum twice each year. We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 & 14 in the spring and 37 & 38 in the fall). The number of simulated tows made reflects true target values for each stratum and sample cells where "tows" are made were randomly selected.

An underlying assumption in all index methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extent of the population. We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct). The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were not sampled for each species. We then use the biomass collected from our samples in contemporary abundance index methods to estimate yearly seasonal population trends.

@@

where $W_i=\frac{S_i}{\sum_{j=1}^N S_j}$ and $y_{k,i}$ represents the biomass in sample $k$ of stratum $i$ in season $s$ and year $y$. These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.

@@

We follow the advice given in [@thorson2019guidance] and [@VASTwiki] to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. We considered the driving factors of movement in our population models as covariates- the dynamic temperature values and static habitat values ($Hab_{J.s}$). We ultimately decided to provide the most information to the model by including exact values for both temperature and habitat covariates to consider a best-case scenario.

@@

Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal]. After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A [@ThorsonChat]. Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). The specifications for Settings A are more  typical for stock assessment application [@ThorsonChat], and we therefore focus our discussion on these VAST results. Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.

Each abundance index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider. The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.

@@

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38). The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain. By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions. As seen in the right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our 20-year abundance estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall absolute errors and standard deviations. Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39. These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

@@

<font size="5">*Abundance Estimate Ratio Results*</font>

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes (Figure \@ref(fig:RatioTrend)).

@@

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.

Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

Including covariate information made a noticeable difference in our Yellowtail model-based VAST estimates. All of the VAST estimates that included covariate information produced a lower absolute error than the corresponding VAST model that did not include covariates. The largest improvements were seen in the increasing temperature scenarios. When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower absolute error than the corresponding stratified mean estimate. This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations. A representative example of Yellowtail abundance estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.

We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling. In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method. Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.

@@

Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values. The different VAST settings produced similar error values with $18/32\approx57\%$ of estimates performing better with VAST Settings A. Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate. VAST produced higher absolute error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates. Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place. These impacts are shown in the plots of representative biomass estimates for Cod shown in Figure \@ref(fig:CodEstEx). Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean. However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.

>>>>

# Estimating Population Trends Under the Pressures of Climate Change

@@

Corresponding author: Ben Levy (blevy6@gmail.com)

@@

Many stock assessments in the United States use biomass estimates derived from stratified random bottom trawl data. To accurately track changes in true biomass of a species, catch observations must contain low noise and the proportion of the population that is sampled should be consistent over time. One increasingly documented phenomenon that violates the latter assumption is climate-induces movement leading to changes in relative biomass between sampled and unsampled portions of the stock area. Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to evaluate two methods of estimating population trend given temperature driven movement. Movement rates combine species-specific static habitat preferences with temperature tolerances. Habitat preferences were derived from niche models relating bottom trawl catches to environmental covariates. A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation. We simulated spatial time series data sets for each species for several temperature scenarios and population trends. Using stratified random sampling on model output we were able to compare biomass estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates. Our focus was on the ability of contemporary index standardization methods to track population trends under shifting spatial preferences.

<!-- We found that in general, both index calculation methods are able to produce similar biomass trends to the model output, with model-based estimates providing more consistent year-to-year estimates. The design-based approach tended to provide biomass estimates that were below the true values in our models, while model-based estimates tended to overestimate true biomass. We highlight scenarios where index calculation methods produced misleading estimates as species shifted their spatial densities over time within their assumed stock boundaries. While including environmental covariates can improve model-based estimates, the outcome is sensitive to the multitude of inputs and settings. The impact of shifting spatial densities became magnified when only a portion of a species’ spatial distribution was sampled. Counter-intuitively, including covariates can sometimes degrade the estimates in this situation. -->

@@

The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf, where depth and latitude define the strata. The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species biomass, and estimating yearly species biomass trends to help inform stock assessments, and ultimately fishing quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each stratum in approximately the same 6 week time period in each season.

One component of a stock assessment is an estimation of yearly biomass that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain biomass estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods typically account for spatial variation in samples by stratifying the sampling area. While it is possible to include environmental influences with design-based methods [@deville1992calibration], it is not a common practice. Model-based biomass estimates use statistical models to measure the relationship between response variables (such as presence or biomass) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process. Common model-based approaches include Generalized Linear Models (GLM), Generalized Additive Model (GAM), and General Linear Mixed Models (GLMM).

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. At the same time a number of recent studies have found a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects]. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate]. A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence]. In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences]. We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of biomass estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.

@@

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depends on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we conduct stratified random sampling (SRS) on model output and analyze the ability of the samples to track population trends. We compare yearly biomass estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean. VAST is a spatial delta-generalized linear mixed model that estimates both biomass and probability of occurrence (presence/absence) [@thorson2019guidance]. VAST also allows users to include covariate data to better inform the model. Covariates can be static (e.g. habitat preferences) or dynamics (e.g. temperature). We explore whether including environmental predictors leads to improved biomass estimates, which is particularly relevant as climate change progresses.

@@

We construct spatial population models for Yellowtail Flounder (*Limanda ferruginea*), Atlantic Cod (*Gadus morhua*), and Haddock (*Melanogrammus aeglefinus*) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock. Movement of each species combines static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct SRS on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50**??** years while Haddock has increased in abundance in recent years [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **just cite 2021 assessments?**. For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To consider a wide range of possibilities, we include increasing, decreasing, and constant population trends for Yellowtail. The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios). Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial data sets.

@@

After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year. We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 & 14 in the spring and 37 & 38 in the fall). The number of simulated tows made reflects true target values for each stratum and sample cells where "tows" are made were randomly selected.

An underlying assumption in all index methods is that individual random samples combine to accurately represent true biomass by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.35. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that SRS will fail to sample the entire geographic extent of the population. We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how the distribution of each species evolved in our increasing temperature scenarios and removing strata where the proportion of a given population was changing over time. Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct). The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were not sampled for each species. We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.

@@

where $W_i=\frac{A_i}{\sum_{j=1}^N A_j}$, $A_i$ is the area of stratum $i$, and $y_{k,i}$ represents the biomass in sample $k$ of stratum $i$ in season $s$ and year $y$. These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.

@@

We follow the advice given in [@thorson2019guidance] and https://github.com/James-Thorson-NOAA/VAST/wiki to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. We considered the driving factors of movement in our population models as covariates- the dynamic temperature values and static habitat values ($Hab_{J.s}$). We ultimately decided to provide the most information to the model by including exact values for both temperature and habitat covariates to consider a best-case scenario.

@@

Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial found at https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/. After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A [@ThorsonChat]. Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). The specifications for Settings A are more  typical for stock assessment application [@ThorsonChat], and we therefore focus our discussion on these VAST results. Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.

Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider. The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.

@@

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38). The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain. By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial data sets with shifting biomass distributions. As seen in the right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our 20-year biomass estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall absolute errors and standard deviations. Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39. These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

@@

<font size="5">*Biomass Estimate Ratio Results*</font>

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined through yearly estimate:model ratio values, where we divide the yearly biomass estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \@ref(fig:RatioTrend)).

@@

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved biomass estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.

Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing biomass estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

Including covariate information made a noticeable difference in our Yellowtail model-based VAST estimates. All of the VAST estimates that included covariate information produced a lower absolute error than the corresponding VAST model that did not include covariates. The largest improvements were seen in the increasing temperature scenarios. When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower absolute error than the corresponding stratified mean estimate. This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations. A representative example of Yellowtail biomass estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.

We see diminished performance of our biomass estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Our analyses have found that the stratified mean tends to under estimate the true biomass and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling. In comparing biomass estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method. Specifically, when comparing the error of biomass estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.

@@

Table \@ref(tab:CodERROR) contains the absolute error between our biomass estimates for Cod and the true model values. The different VAST settings produced similar error values with $18/32\approx57\%$ of estimates performing better with VAST Settings A. Of the biomass estimates without covariates, $12/16=75$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate. VAST produced higher absolute error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates. Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place. These impacts are shown in the plots of representative biomass estimates for Cod shown in Figure \@ref(fig:CodEstEx). Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean. However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.

In comparing biomass estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates. Specifically, when comparing the error of biomass estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates. In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of biomass estimates, decreasing by an average factor of 0.87.

<!-- Similar to Yellowtail results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST biomass still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots? IncTemp fall scenarios)**. -->
